A general Nullstellensatz for generalized spaces

### 2019-03-17 paper-qcoh.tex
Axiom schemes vs. axioms

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\List}{List}
\DeclareMathOperator{\id}{id}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\usepackage{soul}\setul{0.3ex}{}
\let\oldul\ul
\renewcommand{\ul}[1]{\text{\oldul{$#1$}}}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1}}\xspace}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\brak}[1]{\llbracket #1 \rrbracket}
\newcommand{\Mod}[1]{{#1}\mathrm{\text{-}mod}}

@@

\begin{thm}\label{thm:nullstellensatz0}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{equation}\tag{$\ddagger$}\label{eq:nullstellensatz}
\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
\end{equation}
\end{thm}

@@

The statement~\eqref{eq:nullstellensatz} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
substantially broadens the scope of the Nullstellensatz.

@@

Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
it is the universal such source in the sense that any first-order formula
which holds for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz}:

\begin{thm}\label{thm:characterization0}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz}.
\end{enumerate}
\end{thm}

@@

\begin{thm}\label{thm:characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom
scheme~\eqref{eq:nullstellensatz}.
% XXX be more explicit about that being a scheme
\end{enumerate}
\end{thm}

@@

\begin{thm}\label{thm:higher-order-characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
the signature of~$\TT$. Then the following statements are equivalent:
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom scheme~XXX.
\end{enumerate}
\end{thm}

\begin{proof}Theorem~\ref{thm:definability} on the semantic side and the axiom
scheme on the syntactic side allow us to replace any mention of a
powersort~$P(X)$ in~$\alpha$
by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$. Then we
can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
the axiom scheme indeed entails the axiom scheme posited by
Theorem~\ref{thm:characterization}.\end{proof}

>>>>

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\List}{List}
\DeclareMathOperator{\id}{id}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}
\usepackage{soul}\setul{0.3ex}{}
\let\oldul\ul
\renewcommand{\ul}[1]{\text{\oldul{$#1$}}}
\newcommand{\affl}{\ensuremath{{\ul{\AA}^1}}\xspace}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\brak}[1]{\llbracket #1 \rrbracket}
\newcommand{\Mod}[1]{{#1}\mathrm{\text{-}mod}}

@@

\begin{thm}\label{thm:nullstellensatz0}
Let~$\TT$ be a geometric theory. Then, internally to~$\Set[\TT]$:
\begin{equation}\tag{$\dagger$}\label{eq:nullstellensatz0}
\text{A geometric$^*$ sequent~$\sigma$ holds for~$U_\TT$ if and only
if~$\ul{\TT}/U_\TT$ proves$^*$~$\sigma$.}
\end{equation}
\end{thm}

@@

The statement~\eqref{eq:nullstellensatz0} is not a geometric sequent. Therefore
it is not to be expected that it passes from~$\Set[\TT]$ to a
subtopos~$\Set[\TT']$ corresponding to a quotient theory~$\TT'$ of~$\TT$, and
indeed in general it does not. However, there is still a useful substitute,
which we formulate as Theorem~\ref{thm:nullstellensatz-horn}. This substitute
substantially broadens the scope of the Nullstellensatz.

@@

Theorem~\ref{thm:nullstellensatz0} is a source of nongeometric sequents. Indeed,
it is the universal such source in the sense that any first-order formula
which holds for~$U_\TT$ can be deduced from~\eqref{eq:nullstellensatz0}:

\begin{thm}\label{thm:characterization0}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom~\eqref{eq:nullstellensatz0}.
\end{enumerate}
\end{thm}

@@

\begin{thm}\label{thm:characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a first-order
formula over the signature of~$\TT$. Then the following statements are
equivalent.
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in first-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axioms
\begin{equation}\label{eq:nullstellensatz}\tag{$\ddagger$}
  \speak{$\sigma$ holds for~$U_\TT$} \Longrightarrow \speak{$\ul{\TT}/U_\TT$
  proves$^\star$~$\sigma$}
\end{equation}
where~$\sigma$ ranges over the geometric formulas over the signature of~$\TT$.
\end{enumerate}
\end{thm}

% XXX handholding

@@

\begin{thm}\label{thm:higher-order-characterization}
Let~$\TT$ be a geometric theory. Let~$\alpha$ be a higher-order formula over
the signature of~$\TT$. Then the following statements are equivalent:
\begin{enumerate}
\item The formula~$\alpha$ holds for~$U_\TT$.
\item The formula~$\alpha$ is provable in higher-order intuitionistic logic
modulo the axioms of~$\TT$ and the additional axiom scheme
\begin{equation}\label{eq:higher-order-nullstellensatz}\tag{$\P$}
  \speak{the map~$\mathrm{Form}^\star_{\vec
  x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
  \cdots \times X_n)$ is bijective},
\end{equation}
where~$X_1,\ldots,X_n$ is any list of sorts.
\end{enumerate}
\end{thm}

\begin{proof}Theorem~\ref{thm:definability} on the semantic side and the axiom
scheme~\eqref{eq:higher-order-nullstellensatz} on the syntactic side allow us to replace any mention of a
powersort~$P(X)$ in~$\alpha$
by~$\mathrm{Form}^\star_{x\!\?\!X}(\ul{\TT}/U_\TT)/({\dashv\vdash_x})$,
similarly to how the proof of Lemma~\ref{lemma:extended-conservative} compiles
extended geometric logic to ordinary geometric. Then we
can argue as in the proof of Theorem~\ref{thm:characterization}, noting that
the axiom scheme indeed entails the axiom scheme~\eqref{eq:nullstellensatz} posited by
Theorem~\ref{thm:characterization}.\end{proof}
